### 2021-04-26 submission/manuscript.Rmd
Latest draft of the paper after editing the intro and results

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

@@

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice regained most of the weight five days after treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at high levels through thirty days post-infection (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-infection. Surprisingly, PEG-treated mice were still susceptible to *C. difficile* infection after 10-days of recovery from treatment although *C. difficile* was not detectable in most of the group in the initial five days post-infection (Fig. 1C). From 9 days post-infection onward, the median *C. difficile* stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible for up to ten days post-treatment.

@@

#Genera that were different between treatment groups
f_d_group_genera <- read_tsv(here("data/process/5_days_PEG_genus_group_stools.tsv")) %>%
  filter(p.value.adj < 0.05)  #Select only significant p-values
f_d_group_total_genera <- f_d_group_genera %>%
  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
  tally()
#Genera that were different between treatment groups over multiple days
f_d_group_multi_day <- f_d_group_genera %>%
  filter(p.value.adj < 0.05) %>% #Select only significant p-values
  group_by(genus) %>%
  tally() %>%
  filter(n > 1) %>% #select genera that vary over multiple timepoints = n > 1
  arrange(desc(n)) %>%  #Rank by how many time each genera shows up
  nrow() #24 genera includes Peptostreptococcaceae and Unclassified

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015], we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Next, we examined  alpha diversity by looking at Shannon diversity index over time. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X). The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as *Bacteroides* actually decreased in clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria that differ across treatment groups over multiple timepoints. We found `r f_d_group_multi_day` genera were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has a significant impact on the fecal microbiota that was maintained over time and distinct from clindamycin treatment.

Surprisingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine the bacteria that changed during the post-infection period when the group median *C. difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 days post-infection (Fig. S1A). Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C. difficile* was becoming detectable in the stools (Fig. S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X). Although we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

```
**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig. 3A, Data Set S1, Sheet X). Group, day of the experiment, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 (insert code instead) genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and just in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Interestingly, one of the genera that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with sequences that match C. difficile). *Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the timepoint typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-infection because that was when we started to see a large difference in C*. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig. 3E). Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C. difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption** Next, we decided to examine how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered mice either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or a clindamycin treatment before challenging them with *C. difficile 630* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C. difficile* by 7 days post-infection (Fig. 3B). The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity close to baseline by 7 days post-infection (Fig. 3C-D). We found `r recovered_genera` bacteria were impacted by treatment but recovered close to baseline levels within 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, Sheet X). These findings suggest the duration of treatment matters when considering the impact of the laxatives, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. We were also interested in exploring whether PEG might help with engraftment in the context of FMTs. To examine the impact of PEG treatment on *C. difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution via gavage (Fig. 5A). In contrast to our hypothesis, all groups of mice that received PEG actually exhibited prolonged *C. difficile* colonization (Fig. 5B).

* Draft of results

  + Prolonged *C. difficile* colonization in post-CDI PEG treated mice (Fig. 5B)
  + FMT appears to partially restore alpha diversity (Shannon, but not richness Fig. 5C-D)
  + PCoA (necessary?) Could comment on the clustering of Clindamycin and PEG-treated mice that received FMT (Fig. 6A)
  + Only 2 genera significantly impacted by FMT treatment, likely not as important for clearance (Fig. 6B)
  + Bacteria that consistently differ between groups over time, associated with prolonged colonization (Fig. 6C)

@@

```
**Five-day post-infection community data can predict mice that will have prolonged *C. difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the taxa that were influencing prolonged *C. difficile* colonization. We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were colonized with *C. difficile* 10 days post-infection. We chose 5 days post-infection because that was the earliest timepoint where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C. difficile* challenge and then administered an FMT or PBS gavage. The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were driving performance. We selected the top 10 bacteria contributing to our models performance (Fig. 7A) and examined their relative abundance at 5 days post-infection, the timepoint used to predict C. difficile colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that were > 1 % relative abundance in either the cleared or colonized mice and examined the relative abundance dynamics of these bacteria over time. We found _, _, and _ tended to have a higher relative abundance and __, __ had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C).
Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018]. We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is known to be classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020], *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C. difficile* colonization, and had a high relative abundance in the communities of mice that cleared *C. difficile* within 10 days. We identified 4 OTUs that had at least (insert minimum percent identity) to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C. difficile* at 10 days post-infection (Fig. S4).

@@

#Use same code from five_day_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
five_day_mice <- metadata %>% filter(group == "C" & exp_num %in% c("M3","M4", "M5", "M8")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
             group == "WM" & exp_num %in% c("M3","M4", "M5", "M8")|
             group == "WMC" & exp_num %in% c("M3","M4")|
             group == "WMR" & exp_num %in% c("M5","M6"))
five_d_cfu <- five_day_mice %>%
  filter(sample_type == "stool") %>% #Select only stool samples
  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select timepoints on plot
  filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
  group_by(day) %>%
  count()
#Figure out min & max number of samples we had cfu counts for
five_d_cfu_n_min <- min(five_d_cfu$n)
five_d_cfu_n_max <- max(five_d_cfu$n)

@@

#Use same code from one_day_PEG_subset() in code/utilities.R to determine number of samples per day
one_day_samples_per_day <- metadata %>%
  filter(group == "C" & exp_num %in% c("M6")| #Only use C mice from this experiments. Allocated groups to figures based on paper outline.
             group == "1RM1" & exp_num %in% c("M6R")| #Had to differentiate experiment 6 from 6R in the metadata to create unique_mouse_id that wouldn't overlap for the M1 & 1RM1 mice that are both labeled with mouse_ids that are #s1-6
             group == "M1" & exp_num %in% c("M6")) %>%
  group_by(day) %>%
  count()
#To Do: this code is not specific for CFU values or the timepoints shown on panel B
one_day_sample_min <- min(one_day_samples_per_day$n)
one_day_sample_max <- max(one_day_samples_per_day$n)

>>>>

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. However, whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization resistance. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

@@

**5-day laxative treatment leads to prolonged *C. difficile* colonization in mice.** We compared mice treated with the osmotic laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile*, with mice typically clearing *C. difficile* within 10 days post-infection [@Tomkovich2019; @Tomkovich2020]. All PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days, one group was also treated with clindamycin, and one group was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice, with the greatest change in weight observed on the fifth day of PEG treatment and the mice recovered most of the weight five days after treatment (Fig. 1B). After either PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-infection (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-infection. Surprisingly, PEG-treated mice were still susceptible to *C. difficile* infection after 10-days of recovery from treatment, although *C. difficile* was not detectable in most of the group in the initial five days post-infection (Fig. 1C). From 8 days post-infection onward, the median *C. difficile* CFU stabilized for the 10-day recovery group of PEG-treated mice and remained high through 30 days post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible up to ten days post-treatment.

@@

#Genera that were different between treatment groups
f_d_group_genera <- read_tsv(here("data/process/5_days_PEG_genus_group_stools.tsv")) %>%
  filter(p.value.adj < 0.05)  #Select only significant p-values
f_d_group_total_genera <- f_d_group_genera %>%
  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
  tally()
#Genera that were different between treatment groups over multiple days
f_d_group_multi_day <- f_d_group_genera %>%
  filter(p.value.adj < 0.05) %>% #Select only significant p-values
  group_by(genus) %>%
  tally() %>%
  filter(n > 1) %>% #select genera that vary over multiple time points = n > 1
  arrange(desc(n)) %>%  #Rank by how many time each genera shows up
  nrow() #24 genera includes Peptostreptococcaceae and Unclassified

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Next, we examined  alpha diversity by looking at Shannon diversity over time. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X). The majority of the bacterial relative abundances decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment, as *Bacteroides* actually decreased in clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria that differed across treatment groups over multiple time points. We found `r f_d_group_multi_day` genera were different over multiple time points out of the `r f_d_group_total_genera` genera that were different between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine the bacteria that changed during the post-infection period when the group median *C. difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 days post-infection (Fig. S1A). Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C. difficile* was becoming detectable in the stools (Fig. S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

#To Do: Insert code for number of genera different in all 3 types of tissue samples

```
**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the mucosal microbiota between treatment groups at 6 days post-infection with 16 genera that were significantly different in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days post-infection only *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Interestingly, one of the bacteria that was different between treatment groups at 6 days post-infection was *Peptostreptococcaceae* (the genera with an OTU sequence that matches *C. difficile*). *Peptostreptococcaceae* was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-infection because that was when we started to see a large difference in *C. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is twelve (Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**1-day laxative treatment results in transient *C. difficile* colonization and minor microbiota disruption** Next, we decided to examine how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C. difficile* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C. difficile* by 7 days post-infection (Fig. 3B). The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering close to baseline by 7 days post-infection (Fig. 3C-D). We found `r recovered_genera` bacteria were impacted by treatment, but recovered close to baseline levels by 7 days post-infection including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, Sheet X). These findings suggest the duration of treatment matters when considering the impact of the laxatives, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

``` {r Post-CDI PEG stool microbiota}

#Genera that were different between treatment groups
post_group_genera <- read_tsv(here("data/process/post_CDI_PEG_genus_group_stools.tsv")) %>%
  filter(p.value.adj < 0.05)  #Select only significant p-values
post_group_total_genera <- post_group_genera %>%
  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
  tally() #26 total genera

#Genera that were different between treatment groups over multiple days
post_group_multi_day <- post_group_genera %>%
  filter(p.value.adj < 0.05) %>% #Select only significant p-values
  group_by(genus) %>%
  tally() %>%
  filter(n > 1) %>% #select genera that vary over multiple time points = n > 1
  arrange(desc(n)) %>%  #Rank by how many time each genera shows up
  nrow() #24 genera includes Peptostreptococcaceae and Unclassified

```
**Post-CDI laxative treatment disrupts clearance in clindamycin-treated mice regardless of whether an FMT is also administered** Since 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. We were also interested in exploring whether PEG might help with engraftment in the context of FMTs. To examine the impact of PEG treatment on *C. difficile* clearance and FMT treatment, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B). Next we examined how post-CDI PEG treatment impacted the microbiota. Alpha diversity was lower in the PEG-treated mice with the exception of the PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT appeared to partially restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we saw some overlap between the communities of mice that received FMT and the mice treated with only clindamycin after 5 days post-infection (Fig. 6A). The increase in Shannon diversity suggests that the FMT did have an impact on the microbiota, despite seeing prolonged *C. difficile* colonization in the FMT treated mice. However, only *Bacteroidales* and *Porphyromonadaceae* consistently differed between the mice received either an FMT or PBS gavage (Fig. 6B), suggesting the FMT only restored a couple of genera. Overall, we found there were `r post_group_multi_day` bacteria that differed over multiple time points between treatment groups including increased *Akkermansia* and decreased *Ruminococcaceae*, *Clostridiales*, *Lachnospiraceae*, and *Oscillibacter* in mice that received PEG after *C. difficile* challenge (Fig. 6C). In sum, administering PEG actually prolonged *C. difficile* colonization, including in mice that received an FMT, which only restored a couple of bacterial genera.

@@

#To Do: Insert minimum Percent identity as variable name once blast results to Muribaculum have been combined into 1 file

```
**Five-day post-infection community data can predict which mice that will have prolonged *C. difficile* colonization** After identifying bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG treatments, we decided to examine the bacteria that influenced prolonged *C. difficile* colonization. We trained 3 types of machine learning models (random forest, logistic regression, and support vector machine) with input bacterial community data from 5 days post-infection to predict whether the mice were still colonized with *C. difficile* 10 days post-infection. We chose 5 days post-infection because that was the earliest time point where we would see a treatment effect in the mice that were given 1-day PEG treatment three days after *C. difficile* challenge and then administered an FMT or PBS gavage. The random forest model had the highest performance (AUROC = `r max_auc`, Data Set S1, Sheet X), so we next performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C. difficile *colonization. We selected the top 10 bacteria contributing to our models performance (Fig. 7A) and examined their relative abundance at 5 days post-infection, the time point used to predict *C. difficile* colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that had a greater than 1 % relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time. We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C). Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-CDI PEG treated mice.

Previous work examining the impact of PEG on the murine microbiota found that PEG treatment resulted in the permanent loss of S24-7, also known as *Muribaculum intestinale* [@Tropini2018]. We decided to check our *Porphyromonadaceae* OTUs because *Muribaculum intestinale* is classified as *Porphyromonadaceae* by the Ribosomal Database Project (RDP) database [@Vornhagen2020] and *Porphyromonadaceae* was a top feature in the random forest model predicting prolonged *C. difficile* colonization. We identified 4 OTUs that had at least 92% identity to *Muribaculum intestinale* and examined their abundance in mice that either cleared or were still colonized with *C. difficile* at 10 days post-infection. While all of the OTUs, were decreased by PEG and clindamycin treatment, there was some recovery in the mice that cleared (Fig. S4A). We also examined other *Porphyromonadaceae* and *Lachnospiraceae* OTUs since these were the 2 genera that were important to our classification model and contained multiple OTUs that were different at 5 days post-infection between mice that either cleared or remained colonized with *C. difficile* by 10 days post-infection (Data Set S1, sheet X). While individual *Porphyromonadaceae* and *Lachnospiraceae* OTUs tended to be more abundant in the mice that clear *C. difficile* relative to the mice that exhibit prolonged colonization (Fig. S4B-C), there is no single OTU that fits the pattern we observed at the genus level (Fig. 7C), suggesting multiple *Porphyromonadaceae* and *Lachnospiraceae* OTUs influenced *C. difficile* clearance. Overall, our results suggest that specific bacterial community differences explain the prolonged *C. difficile* colonization we observed in 5-day PEG and post-CDI 1-day PEG treated mice.

@@

#Use same code from five_day_PEG_subset() in code/utilities.R to determine number of samples we had C. difficile CFU counts for per day
five_day_mice <- metadata %>% filter(group == "C" & exp_num %in% c("M3","M4", "M5", "M8")| #Only use C mice from these experiments. Allocated groups to figures based on paper outline.
             group == "WM" & exp_num %in% c("M3","M4", "M5", "M8")|
             group == "WMC" & exp_num %in% c("M3","M4")|
             group == "WMR" & exp_num %in% c("M5","M6"))
five_d_cfu <- five_day_mice %>%
  filter(sample_type == "stool") %>% #Select only stool samples
  filter(day %in% c(0, 1, 2, 3, 4, 5, 6, 8, 9, 10, 15, 20, 25, 30)) %>%  #select time points on plot
  filter(!is.na(avg_cfu)) %>% #Remove rows with NA values
  group_by(day) %>%
  count()
#Figure out min & max number of samples we had cfu counts for
five_d_cfu_n_min <- min(five_d_cfu$n)
five_d_cfu_n_max <- max(five_d_cfu$n)

@@

#Use same code from one_day_PEG_subset() in code/utilities.R to determine number of samples per day
one_day_samples_per_day <- metadata %>%
  filter(group == "C" & exp_num %in% c("M6")| #Only use C mice from this experiments. Allocated groups to figures based on paper outline.
             group == "1RM1" & exp_num %in% c("M6R")| #Had to differentiate experiment 6 from 6R in the metadata to create unique_mouse_id that wouldn't overlap for the M1 & 1RM1 mice that are both labeled with mouse_ids that are #s1-6
             group == "M1" & exp_num %in% c("M6")) %>%
  group_by(day) %>%
  count()
#To Do: this code is not specific for CFU values or the time points shown on panel B
one_day_sample_min <- min(one_day_samples_per_day$n)
one_day_sample_max <- max(one_day_samples_per_day$n)
